---
title: Areas of a spatial segregation model significantly different from null expectations
author: ~
date: '2020-05-12'
slug: asymptotic-p-values-for-spatial-segregation-model
categories: ["R"]
tags: ["R Markdown", "point pattern", "R"]
subtitle: ''
summary: ''
authors: []
lastmod: '2020-05-12T8:47:18-04:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", message = FALSE, warning = FALSE)
```

I present code to identify areas of a spatial segregation model that exceed our null expectations using the [relrisk](https://github.com/spatstat/spatstat/blob/master/man/relrisk.Rd) function in the [spatstat](https://github.com/spatstat/spatstat) package and an assumption of normality of the estimated probabilities. A spatial segregation model was originally proposed by [Diggle, Zheng, & Durr in 2005]( https://doi.org/10.1111/j.1467-9876.2005.05373.x) which estimates spatially-varying probabilities of an event of a certain type to occur in an area accounting for other types. The original method uses a Monte Carlo-based simulation, which is computationally expensive. Instead, [Adrian Baddeley](https://github.com/baddstats) and the spatstat team adapted the [relrisk](https://github.com/spatstat/spatstat/blob/master/man/relrisk.Rd) function for a multitype (m > 2) point pattern that has an option to compute the standard error of the probability estimates based on asymptotic theory, assuming a Poisson process. 

Here, I use the standard errors to compute a 95\% confidence interval (CI) at all gridded pixels ("knots") for each type. Knots with a CI that does not capture the null expectation for each type are identified. I use the provided `lansing` dataset from the [spatstat.data](https://github.com/spatstat/spatstat.data) package. Created with assistance from [Dr. Lance Waller](https://sph.emory.edu/faculty/profile/index.php?FID=345) and [Barry Rowlingson](http://barry.rowlingson.com/).

```{r packages}
# Packages
  library(spatstat)
```

```{r data}
# Data
  spatstat.data::lansing

# Convert the lansing data to a ppp object
  ppp_lansing <- spatstat::ppp(x = spatstat.data::lansing$x, 
                             y = spatstat.data::lansing$y,
                             window = spatstat::unit.square(),
                             marks = as.factor(spatstat::marks(spatstat.data::lansing)))

# Plot input
  spatstat::plot.ppp(ppp_lansing, main = "Lansing Woods", cex = 0.5)
```

```{r spatial segregration}
# Estimate nonparametric spatially-varying probabilities of type
  f1 <- spatstat::relrisk.ppp(ppp_lansing, casecontrol = F, diggle = T, se = T, sigma = bw.diggle)
  
# Default plots
  plot(f1$estimate, main = "Probability of an event by type", zlim = c(0,1)) # probabilities
  plot(f1$SE, main = "Standard error of probability", zlim = c(0,0.03)) # standard errors
  
  wh <- spatstat::im.apply(f1$estimate, which.max)
  types <- levels(spatstat::marks(spatstat.data::lansing))
  wh <- spatstat::eval.im(types[wh]) # most common 
  spatstat::plot.im(wh, main="Most common species", ribargs = list(las = 1))
  
# Significant p-values assumming normality of the Poisson process
## relrisk() computes standard errors based on asymptotic theory, assuming a Poisson process
  alpha <- 0.05                           # alpha
  z <- qnorm(alpha/2, lower.tail = F)     # z-statistic
  f1$u <- f1$estimate + z*f1$SE           # Upper CIs
  f1$l <- f1$estimate - z*f1$SE           # Lower CIs
  mu_0 <- as.vector(table(spatstat::marks(ppp_lansing))/ppp_lansing$n) # null expectations by type
  f1$p <- f1$estimate # copy structure of pixels, replace values
  for (i in 1:length(f1$p)) {
    f1$p[[i]]$v <- factor(ifelse(mu_0[i] > f1$u[[i]]$v, "lower",
                                 ifelse( mu_0[i] < f1$l[[i]]$v, "higher", "none")),
                          levels = c("lower", "none", "higher"))
  }

  # Plot significant p-values
  plot(f1$p, main = "Significant difference from null?")
```

The final plot identifies where the spatially-varying probabilities exceed the null expected probability of each type. Areas of higher-than-expected probability could be considered "hot-spots" and areas of lower-than-expected probability could be considered "cold-spots." Areas that are not significantly different from the null expectation could suggest additional sampling is necessary to determine if these areas are hot are cold spots. 
