---
title: Plotting a Neighborhood Network with ggplot2
authors: ["idblr"]
date: '2022-06-26'
slug: neighborhoods-ggplot
categories: ["R"]
tags: ["areal data", "ggplot", "R Markdown", "R"]
subtitle: ''
summary: ''
featured: no
image:
  caption: 'Image credit: [**Ian Buller**](/)'
  focal_point: ''
  preview_only: false
projects: []
output:
  blogdown::html_page:
    pandoc_args: ["--lua-filter=external-links.lua"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", message = FALSE, warning = FALSE)
```

Here is an example of plotting a neighborhoods network with [ggplot2](https://cran.r-project.org/package=ggplot2) using the [sf](https://cran.r-project.org/package=sf) and [sfdep](https://cran.r-project.org/package=sfdep) packages for counties in a U.S. state. 

We can display the "weights" feature of the neighborhoods network as the size of the line segments by scaling the `size` aesthetic with `+ scale_size_identity()`. 

```{r data, results = "hide"}
# Packages
loadedPackages <- c("ggplot2", "sf", "sfdep", "tigris")
invisible(lapply(loadedPackages, require, character.only = TRUE))

# County geometries of Georgia, U.S.A.
shp_ga <- counties(state = "Georgia", cb = TRUE)

# NAD83/UTM zone 17N geospatial projection
proj_ga <- st_transform(shp_ga, crs = 26917)

# First order contiguity (Queen's case by default)
nb <- st_contiguity(st_geometry(proj_ga)) 

# Contiguity-based spatial weights matrix
nbw <- st_weights(nb)

# County centroids
centroids <- st_centroid(proj_ga)

# Assign latitude and longitude for centroid connections in a dataframe
da <- data.frame(
  from = rep(1:length(nbw), attributes(nbw)$comp$d),
  to = unlist(nb),
  weight = unlist(nbw)
)
da <- cbind(
  da, 
  st_coordinates(centroids)[da$from, 1:2], 
  st_coordinates(centroids)[da$to, 1:2]
)
colnames(da)[4:7] <- c("longitude", "latitude", "long_to", "lat_to")
```

Plot counties and first order contiguity line segments with size scaled by "weights" feature.

```{r plot}
ggplot() +
  geom_sf(data = proj_ga, fill = "white", color = "black") + 
  geom_sf(data = centroids, color = "blue", size = 1) + 
  geom_segment(
    data = da,
    aes(
      x = longitude, 
      y = latitude, 
      xend = long_to, 
      yend = lat_to, 
      size = weight
    ),
    color = "red", 
    alpha = 0.5
  ) +
  scale_size_identity() +
  theme_minimal()
```

This answer was also posted to [Stack Overflow](https://stackoverflow.com/a/72763905/6784787). Some code modified from code by @[StupidWolf](https://stackoverflow.com/users/12258459/stupidwolf)'s [answer](https://stackoverflow.com/a/58540394/6784787).
