---
title: Plotting a Neighborhood Network with ggplot2
author: ~
date: '2022-06-26'
slug: neighborhoods-ggplot
categories: ["R"]
tags: ["areal data", "ggplot", "R Markdown", "R"]
subtitle: ''
summary: ''
authors: []
lastmod: '2022-06-26T8:47:18-04:00'
featured: no
image:
  caption: 'Image credit: [**Ian Buller**](/)'
  focal_point: ''
  preview_only: no
projects: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", message = FALSE, warning = FALSE)
```

Here is an example of plotting a neighborhoods network with `ggplot2` using the `sf` and `sfdep` packages for counties in a U.S. state. We can display the "weights" feature of the neighborhoods network as the size of the line segments by scaling the `size` aesthetic with `+ scale_size_identity()`. 

```{r data, results = "hide"}
library(ggplot2)
library(sf)
library(sfdep)
library(tigris)

# County geometries of Georgia, U.S.A.
shp_ga <- counties(state = "Georgia", cb = TRUE)

# NAD83/UTM zone 17N geospatial projection
proj_ga <- st_transform(shp_ga, crs = 26917)

# First order contiguity (Queen's case by default)
nb <- st_contiguity(st_geometry(proj_ga)) 

# Contiguity-based spatial weights matrix
nbw <- st_weights(nb)

# County centroids
centroids <- st_centroid(proj_ga)

# Assign latitude and longitude for centroid connections in a dataframe
da <- data.frame(from = rep(1:length(nbw), attributes(nbw)$comp$d),
                 to = unlist(nb),
                 weight = unlist(nbw))
da <- cbind(da, 
            st_coordinates(centroids)[da$from, 1:2], 
            st_coordinates(centroids)[da$to, 1:2])
colnames(da)[4:7] <- c("longitude", "latitude", "long_to", "lat_to")
```

Plot counties and first order contiguity line segments with size scaled by "weights" feature.

```{r plot}
ggplot() +
  geom_sf(data = proj_ga, fill = "white", color = "black") + 
  geom_sf(data = centroids, color = "blue", size = 1) + 
  geom_segment(data = da,
               aes(x = longitude, y = latitude,
                   xend = long_to, yend = lat_to,
                   size = weight),
               color = "red", alpha = 0.5) +
  scale_size_identity() +
  theme_minimal()
```

This answer was also posted to [Stack Overflow](https://stackoverflow.com/a/72763905/6784787). Some code modified from code by [@StupidWolf](https://stackoverflow.com/users/12258459/stupidwolf)'s [answer](https://stackoverflow.com/a/58540394/6784787)
